---
title: "PolyAIA — worked example"
author: "Yordi Tamayo"
output:
  html_notebook:
    theme: sandstone
    toc: true
    toc_float: true
    fig_width: 12
    fig_height: 5
    fig_caption: true
  html_document:
    theme: sandstone
    toc: yes
    toc_float: yes
    fig_width: 12
    fig_height: 5
editor_options:
  markdown:
    wrap: 72
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, collapse = TRUE)
```

This is a runnable walkthrough of `PolyAIA` on a real mouse cortex
dataset (3xTg / B6, control vs. alcohol; Sanna et al.). It starts from a
**pre-built Seurat object that already carries an annotated `polyA`
assay**, so you can reproduce the interesting part — differential polyA
usage — without needing the raw 10X and polyA-peak files.

Because that object is already complete, the analysis comes first
(sections 1–2). Section 3 then walks through the rest of the workflow:
how the object was built from raw data, and the QC / integration /
visualization helpers.

------------------------------------------------------------------------

# 0. Setup

```{r libraries}
library(PolyAIA)
library(Seurat)
library(dplyr)
```

## Load the example data

The example workspace (`Seurat_PolyA.RData`) contains everything the
walkthrough needs:

| Object | What it is |
|-------------------|-----------------------------------------------------|
| `seurat_polyA` | Seurat object with annotated `RNA` + `polyA` assays |
| `rmsk` | UCSC RepeatMasker track (mm10), used in section 4 |

It is hosted on Mendeley Data and downloaded on first run (a few GB — the
default 60 s timeout is raised, and the file is only fetched if it isn't
already on disk).

```{r download}
# Direct download URL for Seurat_PolyA.RData.
# Get it from the dataset page -> right-click the file -> "Copy link address".
data_url  <- ""                                  # <-- paste the URL here
data_file <- "rds_files/Seurat_PolyA.RData"

dir.create(dirname(data_file), showWarnings = FALSE, recursive = TRUE)

if (!file.exists(data_file)) {
  if (!nzchar(data_url)) {
    stop("Set `data_url` above, or place Seurat_PolyA.RData at: ", data_file)
  }
  options(timeout = 7200)                        # default 60 s is far too short
  download.file(data_url, data_file, mode = "wb")   # mode="wb" is required
}
```

```{r config}
load(file = data_file)
```

The object carries the two assays `PolyAIA` needs, plus the cell-type
annotations produced by Azimuth (`CellType` = broad class,
`Subpopulation` = fine label) and the experimental design columns:

```{r explore}
SeuratObject::Assays(seurat_obj)
table(seurat_obj$CellType)
table(seurat_obj$Strain, seurat_obj$Treatment)
```


------------------------------------------------------------------------

# 1. Differential polyA usage

This is the core of the package. `DEPsMatrix()` folds the whole APA
analysis into one call:

1.  (optional) polyA residuals via `PASTA::CalcPolyAResiduals()` and
    per-comparison differential testing via `DEPolyAPeaks()`;
2.  merging the results with peak- and gene-level metadata;
3.  pseudobulk aggregation;
4.  per-(cell type × comparison) RED / DPAU / gDPAU scoring.

## Define the comparisons

One row per contrast. `ident1`/`ident2` are the identity tokens built
from `condition_cols`; `treatment`/`control` are the pseudobulk
count-column prefixes (they follow the `group.by` order); `Condition` is
the label. Any extra column (here `Strain`, `Treatment`) is stamped onto
every output row.

```{r comparisons}
comparisons <- data.frame(
  ident1    = c("Alcohol_B6",               "Control_3xTg",               "Alcohol_3xTg"),
  ident2    = c("Control_B6",               "Control_B6",                 "Control_B6"),
  treatment = c("B6_Alcohol",               "3xTg_Control",               "3xTg_Alcohol"),
  control   = c("B6_Control",               "B6_Control",                 "B6_Control"),
  Condition = c("B6_Alcohol_vs_B6_Control", "3xTg_Control_vs_B6_Control", "3xTg_Alcohol_vs_B6_Control"),
  Strain    = c("B6",                       "3xTg",                       "3xTg"),
  Treatment = c("Alcohol",                  "Control",                    "Alcohol"),
  stringsAsFactors = FALSE
)
comparisons
```

## Run

```{r deps}
set.seed(100)

deps <- PolyAIA::DEPsMatrix(
  seu             = seurat_obj,
  comparisons     = comparisons,
  celltype_col    = "Subpopulation",
  condition_cols  = c("Treatment", "Strain"),   # -> "Control_B6", "Alcohol_3xTg", ...
  group.by        = c("Strain", "Treatment", "Subpopulation", "Number"),
  gene.names      = "Gene_Symbol",

  # Residual background: Idents built as Treatment+Strain with no separator
  background      = "ControlB6",
  background_cols = c("Treatment", "Strain"),
  background_sep  = "",
  min.counts.background = 5,

  Pasta           = FALSE,        # FALSE = skip residuals/DE, RED from counts only
  min_cells       = 10,          # drop subpopulations with too few cells
  filter_method   = "edgeR",
  repeat_masker   = "Remove",    # drop intronic TE-overlapping polyA sites
  rmsk            = rmsk,
  return_seurat   = FALSE         # also hand back the residual-bearing object
)

red_scores <- deps$red_scores
polyAdb    <- deps$polyAdb   # analyzed peak table (for PolyAPlot)
```

Note: `features` is derived internally from the polyA assay's
`Intron.exon.location` (3'-most exon + intron by default, see
`genomic_positions`), so no separate `Peaksdb()` call is needed.

## The result

```{r inspect}
table(red_scores$Cells, red_scores$Treatment, red_scores$Strain)
```

Each row is one proximal/distal polyA-site pair in one cell type and
comparison. The distal-usage metrics (all treatment − control):

-   **`RED`** — Relative Expression Difference,
    `log2(d/p)_treatment − log2(d/p)_control`.
-   **`DPAU`** — classic 2-site index for that pair,
    `distal / (distal + proximal)` ([Wang et al.
    2022](https://doi.org/10.1073/pnas.2113504118)).
-   **`gDPAU`** — gene-level generalization weighted by site position,
    `1/(n−1) · Σ (i−1)·pᵢ` over all of the gene's sites; equals `DPAU`
    when n = 2.

`RED_Direction` / `RLD_direction` label the shift as `"Lengthened"` or
`"Shortened"`.

```{r top-hits}
red_scores %>%
  filter(!is.na(RED), RED_LRT_padj < 0.05) %>%
  arrange(desc(abs(RED))) %>%
  select(Cells, Condition, Gene_symbol, RED, RED_Direction,
         DPAU, gDPAU, RED_LRT_padj) %>%
  head(15)
```

## Summary plots

A four-panel overview of the RED results: how many polyA sites the analyzed
genes carry, how the tested pairs split between intronic (REDi) and exonic
(REDu) events, and the direction of the significant shifts in one cell type.

```{r summary-plots, fig.width=12, fig.height=9}
library(ggplot2)
library(dplyr)
library(patchwork)

### A. polyA sites per gene ---------------------------------------------------
# One row per gene (npas is gene-level, so distinct() avoids counting a gene
# once per pair/cell type/comparison). Genes with 10+ sites are pooled into "M".
p_npas <- red_scores %>%
  distinct(Gene_symbol, npas) %>%
  mutate(
    npas_bin = ifelse(npas >= 10, "M", as.character(npas)),
    npas_bin = factor(npas_bin, levels = c(as.character(1:9), "M"))
  ) %>%
  count(npas_bin) %>%
  ggplot(aes(x = npas_bin, y = n)) +
  ylim(0,5000) +
  geom_col(fill = "darkorange") +
  geom_text(aes(label = n), vjust = -0.3) +
  labs(x = "poly(A) sites per gene", y = "Gene number")

### B. REDi vs REDu composition -----------------------------------------------
# Pie chart of the tested proximal/distal pairs in L2/3 IT: REDi = the proximal
# site is intronic, REDu = both sites are in the 3' most exon.
p_redtype <- red_scores %>%
  filter(Cells == "L2/3 IT" ) %>%
  count(RED_type) %>%
  ggplot(aes(x = "", y = n, fill = RED_type)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  geom_text(aes(label = paste0(RED_type, "\n(", n, ")")), position = position_stack(vjust = 0.5)) +
  theme_void() +
  theme(legend.position = "none")

### C. REDi distribution (L2/3 IT) --------------------------------------------
# Density of significant intronic events (|RED| > 0.25, padj < 0.1). The median
# annotation shows whether the shift is globally towards lengthening (> 0) or
# shortening (< 0).
df_redi <- red_scores %>%
  filter(Cells == "L2/3 IT" & abs(RED) > 0.25 & RED_LRT_padj < 0.1 & RED_type == "REDi")

p_redi_l23 <- df_redi %>%
  ggplot(aes(x = RED)) +
  xlim(-2,2) +
  ylim(0,0.75) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  annotate("text", x = 0, y = 0.7, label = paste0("median = ", round(median(df_redi$RED), 2))) +
  labs(title = "REDi - L2/3 IT", x = "REDi", y = "Density")

### D. REDu distribution (L2/3 IT) --------------------------------------------
# Same for exonic (3' UTR) events, on identical axes so C and D are directly
# comparable.
df_redu <- red_scores %>%
  filter(Cells == "L2/3 IT" & abs(RED) > 0.25 & RED_LRT_padj < 0.1 & RED_type == "REDu")

p_redu_l23 <- df_redu %>%
  ggplot(aes(x = RED)) +
  xlim(-2,2) +
  ylim(0,0.75) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  annotate("text", x = 0, y = 0.7, label = paste0("median = ", round(median(df_redu$RED), 2))) +
  labs(title = "REDu - L2/3 IT", x = "REDu", y = "Density")

### Assemble ------------------------------------------------------------------
( p_npas | p_redtype ) / ( p_redi_l23 | p_redu_l23) +
  plot_annotation(tag_levels = "A")
```



------------------------------------------------------------------------

# 2. Coverage visualization

`PolyAPlot()` draws coverage tracks with the gene's polyA sites, peaks
and transcript annotation. `cell_filter` selects which cells to show (a
broad class or a list of subpopulations) and `bulk = TRUE` collapses
them into one track; `condition` restricts the plotted region and site
labels to one comparison, and adding `condition_group_cols` also
restricts the coverage cells to the two groups being compared.

```{r polyaplot1, fig.height=8}
PolyAIA::PolyAPlot(
  seu       = seurat_obj,
  gene      = "Mapk10",
  deg_list  = red_scores[red_scores$RED_type == "REDu",],
  polyAdb   = polyAdb,
  species   = "mouse",
  version   = "mm10",
  expand_bp = 2000,
  polyARegion = TRUE,

  cell_filter = "L2/3 IT",            # keep excitatory neurons
  cell_order = c("L2/3 IT"),
  bulk        = TRUE,             # ...as one combined track
  group.by    = "Strain",
  split.by    = "Treatment",

  condition            = "B6_Alcohol_vs_B6_Control",
  condition_group_cols = c("Strain", "Treatment")
)
```

```{r polyaplot2, fig.height=8}
PolyAIA::PolyAPlot(
  seu       = seurat_obj,
  gene      = "Akap11",
  deg_list  = red_scores[red_scores$RED_type == "REDu",],
  polyAdb   = polyAdb,
  species   = "mouse",
  version   = "mm10",
  expand_bp = 2000,
  polyARegion = TRUE,

  cell_filter = "L2/3 IT",            # keep excitatory neurons
  cell_order = c("L2/3 IT"),
  bulk        = TRUE,             # ...as one combined track
  group.by    = "Strain",
  split.by    = "Treatment",

  condition            = "3xTg_Control_vs_B6_Control",
  condition_group_cols = c("Strain", "Treatment")
)
```


```{r polyaplot3, fig.height=8}
PolyAIA::PolyAPlot(
  seu       = seurat_obj,
  gene      = "Msl1",
  deg_list  = red_scores[red_scores$RED_type == "REDu",],
  polyAdb   = polyAdb,
  species   = "mouse",
  version   = "mm10",
  expand_bp = 2000,
  polyARegion = TRUE,

  cell_filter = "L2/3 IT",            # keep excitatory neurons
  cell_order = c("L2/3 IT"),
  bulk        = TRUE,             # ...as one combined track
  group.by    = "Strain",
  split.by    = "Treatment",

  condition            = "3xTg_Alcohol_vs_B6_Control",
  condition_group_cols = c("Strain", "Treatment")
)
```

------------------------------------------------------------------------

# 3. Workflow used to generate Seurat_polyA object

How the example object was built, and the QC / integration /
visualization helpers. The steps that need the raw 10X `.h5` files, polyA
peak counts and fragment files are shown as non-evaluated reference
(`eval = FALSE`); the plotting sections run on the object loaded above.

## 3.1 From 10X to an annotated Seurat object

```{r build-part1, eval=FALSE, warning=FALSE}
seurat_obj <- PolyAIA::UploadSce(
  data_dir        = "../../Mice_example/2_h5_files/",
  file_pattern    = ".h5",
  metadata_file   = "../../Mice_example/4_MetaData/Metada_sanna.csv",
  species         = "mouse",
  reference       = "mousecortexref",   # Azimuth -> CellType / Subpopulation
  column_id       = "Sample_accession",
  genome_build    = "mm10",
  IDtype          = "ENSEMBL",
  remove_doublets = TRUE
)
```

```{r build-part1b, eval=FALSE, warning=FALSE}
# `thresholds` is the per-cell-type table defined in 3.2 below.
seurat_obj <- PolyAIA::SeuratPipeline(seurat_obj, sample_col = "Sample")
```

------------------------------------------------------------------------

## 3.2 Quality control

`QCPlot()` shows the three standard metrics (`nCount_RNA`,
`nFeature_RNA`, `percent_mito`). Passing cutoffs draws them as dashed
reference lines, and `split_by` facets by any metadata column.

```{r qc-plot, fig.height=7, fig.width=13, warning=FALSE}
PolyAIA::QCPlot(seurat_obj, split_by = "CellType")
```

### Per-cell-type filtering

RNA content differs a lot between populations — neurons carry far more
transcripts than microglia or oligodendrocytes — so a single global
cutoff over-filters the low-RNA types and under-filters the high-RNA
ones. `QCFilter()` applies a **different threshold per cell type**: each
cell is matched to its group's row and dropped only if it fails a bound
for *its own* type. Any threshold column you omit simply isn't enforced.

```{r qc-filter}
thresholds <- data.frame(
  CellType     = c("Astro",  "End", "Exc",  "Inh",      "Mic",  "Oligo", "OPC"),
  min_counts   = c( 1000,     400,   2200,   1500,        250,      250,   250),
  max_counts   = c( 5000,    5000,   9000,   9000,       5000,     5000,  5000),
  min_features = c( 1000,     400,   2200,   1500,        250,      250,   250),
  max_features = c( 3000,    2000,   4000,   4000,       2000,     2000,  3000),
  mito_cutoff  = c(    3,       3,      3,      3,          3,        3,     3)
)

seurat_obj <- PolyAIA::QCFilter(seurat_obj, thresholds, group_col = "CellType")
```

```{r qc-plot2, fig.height=7, fig.width=13, warning=FALSE}
PolyAIA::QCPlot(seurat_obj, split_by = "CellType")
```

## 3.3 Cell-type visualization

```{r umap}
PolyAIA::UmapPlot(
  seurat_obj, "harmony",
  color           = "CellType",
  label_by        = "CellType",
  show_labels     = TRUE,
  legend.position = "right"
)
```

`DotPlot()` converts rownames to gene symbols internally (using the
assay's `SYMBOL` feature metadata), so it works whether the object was
built with Ensembl IDs or symbols.

```{r markers}
Gene_type_markers <- list(
  Astro = c("Gfap", "Aqp4", "Aldh1l1", "Gja1", "Slc1a3"),
  Exc   = c("Slc17a7", "Satb2"),
  Inh   = c("Gad1", "Gad2"),
  Mic   = c("C1qa", "C1qb", "C1qc", "Csf1r", "Cx3cr1"),
  Oligo = c("Plp1", "Mbp", "Mobp", "Mog", "Mag"),
  OPC   = c("Pdgfra", "Cspg4", "Sox10", "Olig1", "Olig2")
)

PolyAIA::DotPlot(
  seu       = seurat_obj,
  group_var = "CellType",
  features  = Gene_type_markers,
  dot_scale = 3,
  sizes     = c(7, 0.5, 9, 7)
)
```

------------------------------------------------------------------------

## 3.5 Building and attaching the polyA assay

```{r build-part2, eval=FALSE, warning=FALSE}
library(PASTA); library(Signac); library(EnsDb.Mmusculus.v79)
options(Seurat.object.assay.calcn = TRUE)

counts_dir <- "../../Mice_example/3_polyA_counts/"
samples <- unique(gsub("\\.tab\\.gz$", "", list.files(counts_dir, pattern = "\\.tab\\.gz$")))

# One sample at a time -- keeps peak memory low
polyA.assays <- PolyAIA::UploadPolyAAssays(
  samples       = samples,
  counts_dir    = counts_dir,
  peaks_file    = "../../Mice_example/4_MetaData/polyA_Bam_Files_polyA_peaks.gff",
  fragments_dir = "../../Mice_example/1_Mice_bed_files.nosync/",
  min.features  = 500,
  min.cells     = 50,
  genome        = "mm10"
)

# Unify peaks across samples, then requantify everyone against that set
PeaksFiltering <- PolyAIA::FilterPeaks(polyA.assays, multiple_samples = FALSE)

load("rds_files/Eviroment_seurat.RData")

for (i in seq_along(polyA.assays)) {
  cells <- intersect(colnames(seurat_obj), colnames(polyA.assays[[i]]))
  polyA.assays[[i]] <- subset(polyA.assays[[i]], cells = cells)
}


polyA.unified <- PolyAIA::RequantifyPolyA(
  polyA.assays  = polyA.assays,
  unified.peaks = PeaksFiltering,
  threshold     = 16,
  genome        = "mm10"
)

rm(polyA.assays)

polyA.merged <- merge(polyA.unified[[1]], polyA.unified[-1], merge.data = TRUE)

rm(polyA.unified)

annotations <- Signac::GetGRangesFromEnsDb(ensdb = EnsDb.Mmusculus.v79)
Signac::genome(annotations) <- "mm10"
Signac::Annotation(polyA.merged) <- annotations

cells <- intersect(colnames(seurat_obj), colnames(polyA.merged))
seurat_obj <- subset(seurat_obj, cells = cells)
seurat_obj[["polyA"]] <- subset(polyA.merged, cells = cells)
DefaultAssay(seurat_obj) <- "polyA"

rm(polyA.merged)

# Annotate sites against the polyA reference database
seurat_obj <- PASTA::GetPolyADbAnnotation(
  seurat_obj,
  polyAdb.file = "../../Mice_example/4_MetaData/polyAdb_mm10.txt",
  max.dist     = 50
)

# The example workspace bundles the object together with the RepeatMasker
# table, so section 4 can run straight after load().
rmsk <- read.delim("../../Mice_example/4_MetaData/mm10.rmsk.txt")
```

------------------------------------------------------------------------

# Session info

```{r session-info}
sessionInfo()
```
